---
title: "DyAlloc-vignette"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{DyAlloc-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Disclaimer
Below is the Vignette for my DyAlloc package. The whole package is now on my GitHub and will be submitted on CRAN. This is the link to the package on my GitHub:

https://github.com/ktmiu/DyAlloc/tree/main/Desktop/Cov_project/DyAlloc 

Please refer to it as my second deliverable. Thank you!

# Overview 

 DyAlloc is the implementation of Dynamic Allocation/Covariate Adaptive Randomization. In clinical studies, 
randomized controlled trials are often used to maintain 
unbiased treatment comparison. When assigning subjects to treatment groups, 
complete randomization can avoid selection bias. However, by chance alone, simple randomization has been realized 
to result in imbalances, or unequal  allocation, of treatment 
assignment across study covariates. This is especially true in small trials or in large trials 
with many strata. Given equal variance of treatment responses, this will compromise 
the statistical properties of the study and lead to reduced precision of the treatment 
estimate, and thus increased likelihood of Type II errors (Lebowitsch et al, 2012). 
Therefore, to achieve unbiasedness in subject allocation while maintaining balance 
over covariates, this dynamic allocation scheme has been proposed. 

  Having balanced covariates in clinical trials is crucial. First, it improves the 
credibility of the trial. When it comes to interpreting the analysis results, 
minimization of unbalances during randomization is easier to comprehend than having to adjust 
the analysis after randomization. Second, balance makes conducting interim analysis,
subgroup analysis, or secondary endpoints analysis, etc., much easier. Lastly, 
we are less likely to have power loss due to misspecified models (McEntegart, 2003).

  With the advance of personalized medicine and new-found biomarkers, clinical trials
 often involve multiple centers and an increasingly large number of covariates. This proposes
a challenge to balance on a stratified level while accounting for interactions 
between the covariates. In addition, stratified randomization does not promise overall
balance of the entire study or across different factors. For this reason, minimization 
(dynamic allocation) provides a solution, which balances treatment assignment
over covariate margins. 

# Method
  Dynamic Allocation is an unbalance minimization method proposed by
Lebowitsch et al. They referred to it as the multidimensional dynamic allocation
(MDA) method. Essentially, for each new subject coming into the study, they are assigned a
treatment group based on the existing subjects within each treatment arm. This 
method approaches minimizing unbalances at different levels simultaneously, including
the strata, factors, and overall study. The choice of weight for each level is
decided by the investigators depending on the study design. Furthermore, this method can be
applied to any number of treatment arms.

# Function Descriptions
## 1. Marginal balance as distance measure - measuring imbalance within a population
  A low-level function that computes an imbalance score for each treatment given a set of data.  This function is used primarily used in the subsequent functions and will most probably not be used directly by a practitioner. 
  
  - Function syntax: **MarImb (df, N)**
  - INPUT: This function takes two arguments: 
      - df: the data frame of the study
      - N: the number of treatment arms
  - OUTPUT: a vector of marginal imbalance scores for each treatment arm is produced
    (a total of N scores)
  
## 2. Weighting imbalances - summing imbalances across populations
  This function creates an imbalance score for a new subject to be assigned each possible treatment given data from previous subjects. 

  - Function syntax: **WeiImb (df, N, covars, weight, newsub, site=FALSE)**
  - INPUT: This function takes six arguments: 
      - df: the data frame of the study
      - N: the number of treatment arms
      - covars: a vector of the name of covariates in the study
      - weight: a vector of weights for overall study, with-in stratum, (site), 
        and factors(covariates), in this particular order
      - newsub: a vector of the factor profile for the new subject
      - site: whether or not to account for site imbalance (default:FALSE)
  - OUTPUT: a vector of weighted imbalance scores for each treatment arm is produced
    (a total of N scores)
    
*NOTE: the default for site here is FALSE, but one can specify site=TRUE in the argument
to account for site imbalance. Note that the weight for site has to be placed in the
third element of the 'weight' vector (and can be omitted when site=FALSE)*

## 3. Treatment asssignment probabilities
  This function uses the imbalance scores created from the function WeiImb to produce a vector of probabilities of a new subject being assigned to each treatment group. 
 
  - Function syntax: **Trt_Prob (imbalances, alpha)**
  - INPUT: This function takes two arguments:
      - imbalances: imbalance scores across the N treatment groups 
        (generally, the scores we get from WeiImb would be plugged in here)
      - alpha: the "second best probability" parameter, generally alpha is set to 0.2.
        This means that the treat group a subject assign to gets the probability of 0.8.
  - OUTPUT: a vector of probabilities of being assigned to each treatment group

*NOTE: if the imbalance scores are the same, then they have equal probability of 
being assigned to. E.g. if we get two equal least weighted imbalance scores, they each 
have a probability of 0.5 for a subject to be assigned to.*

## 4. Final treatment assignment - used alone
This function incorporates all the previous functions and gives a final treatment 
group assignment to a new subject. A practitioner can use this function solely
to obtain treatment assignments for the study subjects.

  - Function syntax: **CAR (df, N, covars, weight, newsub, site=FALSE)**
  - INPUT: This function takes six arguments: 
      - df: the data frame of the study
      - N: the number of treatment arms
      - covars: a vector of the name of covariates in the study
      - weight: a vector of weights for overall study, with-in stratum, (site), 
        and factors(covariates), in this particular order
      - newsub: a vector of the factor profile for the new subject
      - site: whether or not to account for site imbalance (default:FALSE)
  - OUTPUT: a treatment group assignment 
  
*NOTE: like the WeiImb function, the site default is FALSE.

# Examples of implementing the functions
```{r setup}
library(DyAlloc)
```
## Without accounting for site imbalance
Here, we considered a clinical trial with 4 covariates: "gender","age","race","trauma severity",
and 4 treatment groups. There are 50 randomized subjects already in the study and we wish to assign
the 51st patient to the least imbalanced group. Produced is the total imbalance score across the 4 
treatment groups and the probability of the 51st patient being assigned to each treatment.
```{r, example without site}
#generate a data frame of 50 subjects in a 4 treatment group study
df1 <- data.frame("gender" = sample(c("female", "male"),50, TRUE),
                 "age" = sample(c("0-35", ">35"),50, TRUE),
                 "race" = sample(c("black", "white"),50,TRUE),
                 "trauma severity"= sample(c("good", "bad"),50,TRUE),
                 "treatment"=sample(c("1","2","3","4"),50, TRUE,c(1,1,5,10)),
                  stringsAsFactors = TRUE)
#covariates of the study
covars <-c("gender","age","race","trauma severity")
#new subject's profile
newsub <-c("male","0-35","black","bad")
#weight of the overall study, the stratum, and the 4 factors 
weight <-c(1,1,rep(1,4))
#print out the final weighted imbalance scores
imb <-WeiImb(df1,4,covars,weight,newsub,FALSE)
imb
#probabilities for the new subject to be assigned to the each of the treatment group 
Trt_Prob(imb,0.2)

#Using CAR function to get the final treatment assignment
DyAlloc(df1,4,covars,weight,newsub,FALSE)
```
The above example shows that the ratio of treatment 1 to 4 already assigned in the study 
is 1:1:5:10. Given equal weights to each treatment arm, we can see that the probability
of the 51st subject's assignment are most likely to fall into treatment 1 or 2 with a 50/50 chance, which are the least imbalanced ones.

## Accounting for site imbalance 
The same data frame but also accounting for site in the total imbalance score.
```{r, example with site}
df <- data.frame("gender" = sample(c("female", "male"),50, TRUE),
                 "age" = sample(c("0-35", ">35"), 50, TRUE),
                 "race" = sample(c("black", "white"),50,TRUE),
                 "trauma severity"= sample(c("good", "bad"),50,TRUE),
                 "treatment"=sample(c("1","2","3","4"),50, TRUE),
                 "site"=sample(c("1","2"),50, TRUE),
                 stringsAsFactors = TRUE)
covars <-c("gender","age","race","trauma severity")
#new subject's profile with the first element being site
newsub<-c("2","male","0-35","black","bad")
#weight of the overall study, the stratum, the site, and the 4 factors 
weight <-c(1,1,1,rep(1,4))
imb<-WeiImb(df,4,covars,weight,newsub,TRUE)
imb
Trt_Prob(imb,0.2)
DyAlloc(df,4,covars,weight,newsub,TRUE)
```
This example shows that when there is random treatment assignment in the study,
given equal weights to each treatment arm, the 51st subject's assignemnt can also
be random. Under the "second best probability" paradigm, the final treatment assignment
would most likely fall into the treatment arm with a probability of 0.8.

# Simulation 

200 Subject profiles were simulated with 4 covariates considered - gender, age,
race, and trauma severity, in a 4-treatment-arms clinical trial. Site imbalance 
is not considered in this scenario. Each subject is assigned to a treatment group
after the simulation using two methods:  covariate adaptive randomization and simple randomization.
To make this more interesting and to see the effect of balanced
allocation, the ratio of female to male is set to be 1:4, and the ratio of Black to White is 7:3. 

It can be seen that the covariate adaptive randomization did a very good job in balancing treatment assignment across individual variables as well as across strata defined by interactions between variables, especially compared to simple randomization.  For example, no White females were asigned to treatment 1 using simple randomization, while exactly 25% of White females were assigned to treatment 1 using covaraite adaptive randomization, which is the target proportion. 

```{r, include=FALSE}
set.seed(12345)
mydf <- data.frame("gender" = sample(c("female", "male"),200, TRUE,c(.2,.8)),
                   "age" = sample(c("0-35", ">35"),200, TRUE),
                   "race" = sample(c("black", "white"),200,TRUE,c(0.7,0.3)),
                   "trauma severity"= sample(c("good", "bad"),200,TRUE), 
                   stringsAsFactors = TRUE)
covars <-c("gender","age","race","trauma severity")
weight <-c(1,1,rep(1,4))
treatment <- vector()
newdf <- data.frame("gender"=NA,"age"=NA,"race"=NA,"trauma severity"=NA,"treatment"=NA)
for (i in 1:200){
  newsub<-c(as.character(mydf[i,1]),as.character(mydf[i,2]),as.character(mydf[i,3]),
            as.character(mydf[i,4]))
  newtreatment <- DyAlloc(newdf,4,covars,weight,newsub,FALSE)
  treatment <- append(treatment,newtreatment)
  newdf <- cbind(mydf[1:i,],treatment)
}
newdf
newdf$srs = sample(c("1", "2", "3", "4"), 200, replace = TRUE)
```
i) Table content for the number of female vs. male
ii) Table for the proportion of each treatment assignment in female and in male using the covariate adaptive randomization
iii) Table for the proportion of each treatment assignment in female and in male using simple randomization
```{r, echo=FALSE}
#number of female and male in the newdf
table(newdf$gender)
#proportion of each gender in each treatment 
prop.table(table(newdf$gender,newdf$treatment),1)
#proportion of each gender in each treatment 
prop.table(table(newdf$gender,newdf$srs),1)
```
i) Table content for the number of Black vs. White
ii) Table for the proportion of each treatment assignment in Black and in White  using the covariate adaptive randomization
iii) Table for the proportion of each treatment assignment in Black and in White using simple randomization 
```{r, echo=FALSE}
#disproportional race example
table(newdf$race)
prop.table(table(newdf$race,newdf$treatment),1)
prop.table(table(newdf$race,newdf$srs),1)
```

i) Table content for the number in each of the 4 combinations of gender and race
ii) Table for the proportion of each treatment assignment in each of the 4 combinations of gender and race using the covariate adaptive randomization
iii) Table for the proportion of each treatment assignment in each of the 4 combinations of gender and race  using simple randomization 
```{r, echo=FALSE}
#disproportional race example
table(newdf$race, newdf$gender)

ctab = rbind(prop.table(table(newdf$race[newdf$gender=="female"],newdf$treatment[newdf$gender=="female"]),1), prop.table(table(newdf$race[newdf$gender=="male"],newdf$treatment[newdf$gender=="male"]),1))
dimnames(ctab)[[1]] = c("black female", "white female", "black male", "white male")
ctab


ctab.srs = rbind(prop.table(table(newdf$race[newdf$gender=="female"],newdf$srs[newdf$gender=="female"]),1), prop.table(table(newdf$race[newdf$gender=="male"],newdf$srs[newdf$gender=="male"]),1))
dimnames(ctab.srs)[[1]] = c("black female", "white female", "black male", "white male")
ctab.srs
```
